---
title: "ps4_yeomans"
author: "Sydney Yeomans"
format:
  html:
    embed-resources: true
editor: visual
---

## Tidyverse packages

```{r}
library(tidyverse)
library(dplyr)

```

## Problem 1

Use the nzge data for this problem. Load the data set

```{r}
nzge <- load("~/R and Git/STATS-506-PS4/nzge.rda")
```

a. Generate a table (which can just be a nicely printed tibble) of vote count (regardless of party) per year/type. Make sure to sort it by vote count.

```{r}
vote_count <- nzge %>% 
            group_by(voting_type, election_year) %>% 
            summarize(count = sum(votes)) %>%
            arrange(desc(count)) %>%
            ungroup()
vote_count
```

b. Focus only on the 2014 election. Report the proportion of votes for each party in the Candidate election. Again, produce a nice table and sort it by percent of vote.

```{r}
year_2014 <- nzge %>% 
            filter(election_year == "2014", voting_type == "Candidate") %>%
            group_by(party) %>%
            summarise(votes = sum(votes)) %>%
            mutate(proportion_votes = (votes / sum(votes)) * 100) %>%
            arrange(desc(proportion_votes)) %>%
            ungroup()
year_2014

```

c. Produce a nice table indicating, for each year, which party won the Candidate vote and which party won the Party vote.

```{r}
who_won_what <- nzge %>%
                group_by(election_year, voting_type, party) %>%
                summarize(total_votes = sum(votes), .groups = "drop") %>%
                group_by(election_year, voting_type) %>%
                mutate(max_votes = max(total_votes)) %>%
                filter(total_votes == max_votes) %>%
                arrange(election_year, voting_type) %>%
                ungroup()
who_won_what

```



## Problem 2

Use the “ATP Matches” data from 2019

a.

```{r}

```

b.

```{r}

```

c.

```{r}

```

d.

```{r}

```





## Problem 3

Use the NYTimes Covid data

a.

```{r}

```

b.

```{r}

```

c.

```{r}

```



